\chapter{Model Scaling and Efficiency}

Every architectural finding in Chapters 4--7 was established at a fixed
model size. This chapter asks whether those findings hold at other
scales, and where on the performance-cost curve a practitioner should
operate. A secondary question is whether scaling behaviour depends on
task difficulty: a 1v1 discrimination problem and a multiclass problem
may saturate at different scales for physically meaningful reasons.

All experiments use a common fixed baseline configuration: T1 =
\texttt{identity}, best D01 from Ch.\,4, E1 = \texttt{sinusoidal}, no
physics biases. This isolates scaling as the sole variable.
Model capacity is controlled via the convenience key
\texttt{axes/model\_size\_key}, which bundles dim, depth, heads, and
FFN dim into named presets spanning roughly 50k to 4M parameters.

% ──────────────────────────────────────────────
\section{Scaling curves across tasks (Exp.\ 8A)}
\textit{Lens: performance and efficiency.}

Trains all four size presets on three classification tasks. The central
question is whether the saturation point, the model size beyond which
additional parameters yield diminishing returns, shifts with task
difficulty. A steeper and later-saturating curve on the harder 1v1
tasks would reflect a genuine physics difference, not a modelling
artefact.

\begin{table}[h]
\centering\small
\begin{tabular}{llr}
\toprule
Axis & Values & Config \\
\midrule
Model size & \texttt{d32\_L3}, \texttt{d64\_L6} & 4 \\
           & \texttt{d128\_L8}, \texttt{d256\_L12} & \\
Task       & 4t vs.\ background & 3 \\
           & 4t vs.\ ttH & \\
           & multiclass (all five processes) & \\
Seeds & & 3 \\
\midrule
Total & & \textbf{36} \\
\bottomrule
\end{tabular}
\caption{Exp.\ 8A. Common baseline configuration throughout.}
\end{table}

Outputs: three scaling curves (one per task) on the same axes, showing
seed-averaged AUROC vs.\ parameter count on a log scale. A secondary
panel shows signal efficiency at fixed background rejection $1/B = 50$
as a physics-operational complement to AUROC.

% ──────────────────────────────────────────────
\section{Efficiency metrics (Exp.\ 8B)}

No new training required. For each checkpoint from Exp.\ 8A, three
efficiency metrics are recorded: inference throughput in events per
second, training wall-clock time per epoch, and epochs to 95\% of
final AUROC. These are plotted as a Pareto panel against AUROC,
one curve per task. The model sitting on the Pareto frontier at each
task is the practical recommendation for that setting.

% ──────────────────────────────────────────────
\section{Axis transfer across scales (Exp.\ 8C)}
\textit{Primary task only.}

A targeted check that the axis rankings established in Chapters 4--7
are stable across scales. The two most impactful axes from Ch.\,8
are trained at the smallest and largest presets in addition to the
default, using the primary task only. Scale-dependent conclusions
would imply the Part II findings are specific to the medium-size
regime; scale-independent conclusions support them as broadly
applicable.

\begin{table}[h]
\centering\small
\begin{tabular}{llr}
\toprule
Axis & Values & Config \\
\midrule
Top-2 axes from Ch.\,8 & best vs.\ default setting & 2 \\
Model size & \texttt{d32\_L3}, \texttt{d64\_L6}, \texttt{d256\_L12} & 3 \\
Seeds & & 3 \\
\midrule
Total & & \textbf{18} \\
\bottomrule
\end{tabular}
\caption{Exp.\ 8C. Primary task (4t vs.\ background) only.
Top-2 axes determined after Ch.\,8 analysis is complete.}
\end{table}

\subsection*{Chapter summary}

\begin{table}[h]
\centering\small
\begin{tabular}{llr}
\toprule
Experiment & Description & Runs \\
\midrule
8A — Scaling curves  & 4 sizes $\times$ 3 tasks $\times$ 3 seeds & 36 \\
8B — Efficiency      & post-hoc from 8A checkpoints              &  0 \\
8C — Axis transfer   & 2 axes $\times$ 3 sizes $\times$ 3 seeds  & 18 \\
\midrule
\textbf{Total} & & \textbf{54} \\
\bottomrule
\end{tabular}
\end{table}
