\chapter{The Best Model and Its Physics Reach}

This chapter does two things. First, it assembles the recommended
configuration from the Ch.\,8 analysis, trains it at the Pareto-optimal
scale from Ch.\,9, and compares it against a simple baseline and the
external MIParT benchmark. Second, it attempts to translate the AUROC
improvement into a physics statement: how much does a better classifier
actually matter in units of LHC luminosity?

% ──────────────────────────────────────────────
\section{The recommended model (Exp.\ 10A)}

The recommended configuration is taken directly from the Ch.\,8
surrogate analysis. It is trained at the Pareto-optimal model size
from Ch.\,9 using five seeds to reduce Monte Carlo uncertainty on the
final performance estimate. Two baselines are included for comparison:
the simplest possible transformer (T1=\texttt{raw}, no biases, smallest
preset) and MIParT (Li et al., 2024) evaluated on the same dataset and
task.

\begin{table}[h]
\centering\small
\begin{tabular}{llr}
\toprule
Model & Description & Runs \\
\midrule
Recommended & best config from Ch.\,8, Pareto size from Ch.\,9 & 5 \\
Simple baseline & T1=\texttt{raw}, no biases, \texttt{d32\_L3} & 3 \\
MIParT reference & Li et al.\ on same dataset & 3 \\
\midrule
Total & & \textbf{11} \\
\bottomrule
\end{tabular}
\caption{Exp.\ 10A. Both tasks (4t vs.\ background and 4t vs.\ ttH)
are evaluated for all three models.}
\end{table}

Outputs: AUROC and signal efficiency at $1/B = 50$ for all three
models on both tasks, with seed-averaged means and error bars. This
is the thesis headline number.

% ──────────────────────────────────────────────
\section{From AUROC to physics sensitivity (Exp.\ 10B)}

\textit{Note: this section describes machinery I am still working
through and may not appear in the final thesis, depending on whether
the results are meaningful and the methodology is sufficiently under
control. It is included here as a planning placeholder.}

The quantity a physics analysis actually cares about is not AUROC but
sensitivity to the four-top-quark production cross-section. The
connection is made through a simplified profile likelihood: given the
classifier score for each event, a binned discriminant is constructed,
signal and background yields are scaled to a target luminosity, and
the expected significance is computed under the asymptotic
approximation of Cowan et al.\ (2011). In the absence of systematics
this reduces to the familiar form

\begin{equation}
Z \approx \frac{S}{\sqrt{B}},
\end{equation}

where $S$ and $B$ are the expected signal and background yields at
the chosen working point.

The main output is a luminosity-versus-significance curve for the
recommended model, the simple baseline, and MIParT. The headline
number is $\mathcal{L}_{5\sigma}$, the luminosity at which each
classifier first reaches $Z = 5$. A reduction in $\mathcal{L}_{5\sigma}$
directly quantifies how much LHC running time the architectural
improvements of this thesis are worth.

All numbers are relative comparisons between classifiers at the same
luminosity assumption, using MC yields only with no systematic
uncertainties. They should not be read as absolute physics predictions.
